---
title: "Palo Verde Electricity Hub Analysis (2014-2024)"
author: "Kendall Beaver"
date: "2024-06-21"
output:
  pdf_document: default
  html_document: default
---

# Overview

The [Intercontinental Exchange (ICE)](https://www.theice.com/index) has provided the U.S. Energy Information Administration selected wholesale daily spot price data from 2001-2024 (<https://www.eia.gov/electricity/wholesale/>) for 8 electricity hubs:

1.  ERCOT North 345KV Peak

2.  Indiana Rt Peak

3.  Mid C Peak

4.  Nepool MH DA LMP Peak

5.  NP15 EZ Gen DA LMP Peak

6.  Palo Verde Peak

7.  PJM WH Real Time Peak

8.  SP-15 Gen DA LMP Peak

This report has analyzed the past 10 years of daily spot price data for Palo Verde Peak and discovered that trades are affected primarily by change in temperature, energy consumption, but in one rare case, a PPA.

# Code

```{r setup, message=FALSE, warning=FALSE, results='hide'}
# Import libraries
library(readxl)
library(dplyr)
library(ggplot2)
```

```{r import weather dataset, message=FALSE, warning=FALSE, results='hide'}
# Import weather data
weather_data <- read.csv ("./Data/phx_weather_2021.csv")
```

```{r 2014-2015, message=FALSE, warning=FALSE, results='hide'}

data_2014 <- read_excel ("./Data/ice_electric-2014final.xls")

# Need to correct "Nepool MH Da LMP Peak" to "Nepool MH DA LMP Peak"
# Need to correct "NP 15 EZ..." to "NP15 EZ Gen..."
# Correct "Palo Verde" to "Palo Verde Peak"
# Change "SP-15 Gen DA LMP Peak" to "SP15 EZ Gen DA LMP Peak"
colnames(data_2014) <- gsub("\n", " ", colnames(data_2014))
colnames(data_2014) <- gsub("  ", " ", colnames(data_2014))

data_2014 <- data_2014 %>%
  mutate(`Price hub` = 
           ifelse(`Price hub` == "Nepool MH Da LMP Peak", "Nepool MH DA LMP Peak",
           ifelse(`Price hub` == "Indiana Hub RT Peak", "Indiana Rt Peak",
           ifelse(`Price hub` == "PJM-Wh Real Time Peak", "PJM WH Real Time Peak",
           ifelse(`Price hub` == "Mid Columbia Peak", "Mid C Peak",
           ifelse(`Price hub` == "PJM Wh Real Time Peak", "PJM WH Real Time Peak",
           ifelse(`Price hub` == "NP 15 EZ Gen DA LMP Peak", "NP15 EZ Gen DA LMP Peak",
           ifelse(`Price hub` == "Palo Verde", "Palo Verde Peak",
           ifelse(`Price hub` == "SP15 EZ Gen DA LMP Peak", "SP-15 Gen DA LMP Peak", `Price hub`)))))))))

data_2015 <- read_excel ("./Data/ice_electric-2015final.xls")
colnames(data_2015) <- gsub("\n", " ", colnames(data_2015))
colnames(data_2015) <- gsub("  ", " ", colnames(data_2015))
unique(c(colnames(data_2015)))
data_2015

data_2016 <- read_excel ("./Data/ice_electric-2016final.xls")
colnames(data_2016) <- gsub("\n", " ", colnames(data_2016))
colnames(data_2016) <- gsub("  ", " ", colnames(data_2016))
unique(c(colnames(data_2016)))
data_2016

data_2017 <- read_excel ("./Data/ice_electric-2017final.xlsx")
colnames(data_2017) <- gsub("\r\n", " ", colnames(data_2017))
colnames(data_2017) <- gsub("  ", " ", colnames(data_2017))
unique(c(colnames(data_2017)))
data_2017

data_2018 <- read_excel ("./Data/ice_electric-2018final.xlsx")
colnames(data_2018) <- gsub("\r\n", " ", colnames(data_2018))
colnames(data_2018) <- gsub("  ", " ", colnames(data_2018))
unique(c(colnames(data_2018)))
data_2018

data_2019 <- read_excel ("./Data/ice_electric-2019final.xlsx")
colnames(data_2019) <- gsub("\r\n", " ", colnames(data_2019))
colnames(data_2019) <- gsub("  ", " ", colnames(data_2019))
unique(c(colnames(data_2019)))
data_2019

data_2020 <- read_excel ("./Data/ice_electric-2020final.xlsx")
colnames(data_2020) <- gsub("\r\n", " ", colnames(data_2020))
colnames(data_2020) <- gsub("  ", " ", colnames(data_2020))
unique(c(colnames(data_2020)))
data_2020

data_2021 <- read_excel ("./Data/ice_electric-2021final.xlsx")
colnames(data_2021) <- gsub("\r\n", " ", colnames(data_2021))
colnames(data_2021) <- gsub("  ", " ", colnames(data_2021))
unique(c(colnames(data_2021)))
data_2021

data_2022 <- read_excel ("./Data/ice_electric-2022final.xlsx")
data_2022 <- select(data_2022, -'Trade date') # REMOVE COLUMN because it's bad
data_2022$'Trade date' <- data_2022$`Delivery start date` - 86400
data_2022 <- data_2022 %>% select(1, `Trade date`, everything()) # Make 'Trade date' the 2nd column like all the other datasets
colnames(data_2022) <- gsub("\r\n", " ", colnames(data_2022))
colnames(data_2022) <- gsub("  ", " ", colnames(data_2022))
unique(c(colnames(data_2022)))
data_2022

data_2023 <- read_excel ("./Data/ice_electric-2023final.xlsx")
colnames(data_2023) <- gsub("\r\n", " ", colnames(data_2023))
colnames(data_2023) <- gsub("  ", " ", colnames(data_2023))
unique(c(colnames(data_2023)))
data_2023

data_2024 <- read_excel ("./Data/ice_electric-2024.xlsx")
colnames(data_2024) <- gsub("\r\n", " ", colnames(data_2024))
colnames(data_2024) <- gsub("  ", " ", colnames(data_2024))
unique(c(colnames(data_2024)))
data_2024
```

## Head 2014-2024

Use the following to make mass changes:

```{r head, message=FALSE, warning=FALSE, results='hide'}

# 8 Peaks (2014-2019):
#
# "ERCOT North 345KV Peak"  
# "Indiana Rt Peak"         
# "Mid C Peak"              
# "Nepool MH DA LMP Peak"   
# "NP15 EZ Gen DA LMP Peak"
# "Palo Verde Peak"         
# "PJM WH Real Time Peak"
# "SP-15 Gen DA LMP Peak"

unique(data_2014$`Price hub`)
unique(data_2015$`Price hub`) 
unique(data_2016$`Price hub`)
unique(data_2017$`Price hub`)
unique(data_2018$`Price hub`)
unique(data_2019$`Price hub`)

# 7 Peaks (2020-2024):

# Indiana Hub RT Peak
# Mid C Peak
# Nepool MH DA LMP Peak
# NP15 EZ Gen DA LMP Peak
# Palo Verde Peak
# PJM WH Real Time Peak
# SP15 EZ Gen DA LMP Peak
# 
# Missing is "ERCOT North 345KV Peak"

unique(data_2020$`Price hub`) # 7 Peaks
unique(data_2021$`Price hub`) # 7 Peaks
unique(data_2022$`Price hub`) # 7 Peaks
unique(data_2023$`Price hub`) # 7 Peaks
unique(data_2024$`Price hub`) # 7 Peaks
```

## Combined Dataset

```{r combined dataset, message=FALSE, warning=FALSE, results='hide'}

combined_data <- rbind(data_2014, data_2015, data_2016, data_2017, data_2018, data_2019, data_2020, data_2021, data_2022, data_2023, data_2024)
combined_data
```

```{r plots, message=FALSE, warning=FALSE, results='hide'}

# Plot the data

# Make sure the dates are in the correct format
combined_data$`Trade date` <- as.Date(combined_data$`Trade date`)

# Filter dataset for "Palo Verde Peak"
palo_verde_peak <- combined_data %>% filter(`Price hub` == 'Palo Verde Peak')

# Rearrange "Change" column from largest to smallest
palo_verde_peak %>% arrange(desc(Change))

# Look at top results
head(palo_verde_peak)
# Top 6 trade dates are:
# 2021-06-16
# 2020-08-17
# 2023-08-15
# 2022-09-05
# 2022-08-31	
# 2020-09-03

# Let's look at June 2021:
pvp_june_2021 <- palo_verde_peak %>%
  filter(`Trade date` > as.Date('2021-06-01') & `Trade date` < as.Date('2021-06-30'))
pvp_june_2021

ggplot(pvp_june_2021, aes(x = `Trade date`, y = Change)) +
  geom_line() +
  labs(title = "Time Series Line Graph - June 2021",
       x = "Change in Price",
       y = "Value") +
  theme_minimal()

# Let's look at August & September 2020:
pvp_aug_2020 <- palo_verde_peak %>%
  filter(`Trade date` > as.Date('2020-08-01') & `Trade date` < as.Date('2020-08-31'))
pvp_aug_2020

pvp_sept_2020 <- palo_verde_peak %>%
  filter(`Trade date` > as.Date('2020-09-01') & `Trade date` < as.Date('2020-09-30'))
pvp_sept_2020
```

Some additional factors to look at:

-   \$84 million in higher wholesale short-term sales primarily due to: (i) an increase in price and sales volume; and (ii) an increase in capacity sales to affiliates for a tolling PPA (Power Purchase Agreement) entered into in June 2021; (<https://www.sec.gov/ix?doc=/Archives/edgar/data/100122/000010012222000020/tep-20211231.htm>)

## Import Weather Dataset

## Fixing Hidden Characters in Columns

```{r hidden characters, message=FALSE, warning=FALSE, results='hide'}

# This showed me that there were hidden columns

unique(c(colnames(data_2014))) # "Delivery \nend date"
unique(c(colnames(data_2015))) # "Delivery \nend date"
unique(c(colnames(data_2016))) # "Delivery \nend date"
unique(c(colnames(data_2017))) # "Delivery \r\nend date"
unique(c(colnames(data_2018))) # "Delivery \r\nend date"
unique(c(colnames(data_2019))) # "Delivery \r\nend date"
unique(c(colnames(data_2020))) # "Delivery \r\nend date"
unique(c(colnames(data_2021))) # "Delivery \r\nend date"
unique(c(colnames(data_2022))) # "Delivery \r\nend date"
unique(c(colnames(data_2023))) # "Delivery \r\nend date"
unique(c(colnames(data_2024))) # "Delivery \r\nend date"
```
